Full Dissertation

### 2024-04-11 Sections/hold_temp.qmd
finish intro

@@

@@

>>>>

A consequence of varied training is that the trial to trial sequences will tend to to consist of fewer repeats, and more switches between task condition (i.e. throwing locations; category exemplars; target velocitys), or in the case where variability is being manipulated by the dispersion of training items, the trial to trial sequences of varied participants will tend to consist of larger jumps in the stimulus space.

 The number of switches can vary greatly, with the two extremes being varied participants completing all of their training trials in one block before switching to the next condition (blocked sequencing), or if they alternate between conditions on a trial by trial basis (random/intermixed/interleaved sequencing).

  [@sheaContextualInterferenceEffects1979]. In this seminal study, both groups of training subjects trained with the same number of trials of three separate movement patterns. A blocked group that completed all of their trials with one sequence before beginning the next sequence, and a random group that trained with all three movement patterns interspersed throughout the course of training. Participants were also randomly assigned to retention testing under either blocked or random sequence conditions, thus resulting in all four training-testing combinations of blocked-blocked; blocked-random; random-blocked; random-random. There was some effect of sequence context, such that both groups performed better when the testing sequence matched their training sequence. However, the main finding of interest was the advantage of random-training, which resulted in superior testing performance than blocked training regardless of whether the testing stage had a blocked or random sequence, an effect observed both immediately after training, and in a follow up test ten days after the end of training. A more recent study [@chuaPracticeVariabilityPromotes2019], found a similar pattern of results in both a varied vs. constant training manipulation (expt 1, bean-bag throwing task), and of random training vs. blocked training manipulation (expt 2 & 3, bean-bag throwing & golf putting). The novelty of this study is that the experimenters queried subjects about their attentional focus throughout the training stage. In all three experiments varied or random trained-subjects reported significantly greater external attention (e.g. attending to the target distance), and constant or blocked subjects reported more internal attention (e.g. posture or hand position). The authors argue that the benefits of varied/random training may thusly be mediated by changes in attentional focus.

### Sequence Effects

A necessary consequence of varied training is that participants will have the experience of switching from one task condition to another. The number of switches can vary greatly, with the two extremes being varied participants completing all of their training trials in one before switching to the next condition (blocked sequencing), or if they alternate between conditions on a trial by trial basis (random/intermixed/interleaved sequencing). Not long after the initial influx of schema-theory inspired studies testing the benefits of variability hypothesis was shown that the influence of varied training might interact with the type of training sequence chosen by the experimenter  [@sheaContextualInterferenceEffects1979]. In this seminal study, both groups of training subjects trained with the same number of trials of three separate movement patterns. A blocked group that completed all of their trials with one sequence before beginning the next sequence, and a random group that trained with all three movement patterns interspersed throughout the course of training. Participants were also randomly assigned to retention testing under either blocked or random sequence conditions, thus resulting in all four training-testing combinations of blocked-blocked; blocked-random; random-blocked; random-random. There was some effect of sequence context, such that both groups performed better when the testing sequence matched their training sequence. However, the main finding of interest was the advantage of random-training, which resulted in superior testing performance than blocked training regardless of whether the testing stage had a blocked or random sequence, an effect observed both immediately after training, and in a follow up test ten days after the end of training.

Prior to the influential @sheaContextualInterferenceEffects1979 study, studies investigating the benefits of variability hypothesis had utilized both blocked and random training schedules, often without comment or justification. It was later observed [@leeInfluencePracticeSchedule1985] that positive evidence for benefits of varied training seemed more likely to occur for studies that utilized random schedules. The theoretical basis of such studies was invariably an appeal to Schmidt's schema theory; however schema theory made no clear predictions of an effect of study sequence on retention or generalization, thus prompting the need for alternate accounts. One such account, the elaborative processing account [@sheaContextEffectsMemory1983], draws on the earlier work [@battigFacilitationInterference1966] and argues that randomly sequencing conditions during training promotes comparison and contrastive processes between those conditions, which result in a deeper understanding of the training task than could arise via blocked sequencing. Supporting evidence for elaborative processing comes in the form of random-sequence trained subjects self-reporting more nuanced mental representations of movement patterns following training [@sheaContextEffectsMemory1983], and by manipulating whether subjects are able to perform comparisons during training [@wrightContributionElaborativeProcessing1992]. An alternative, though not incompatible account suggests that the benefits of random-sequencing are a result of such sequences forcing the learner to continually reconstruct the relevant motor task in working memory [@leeLocusContextualInterference]. Blocked training, on the other hand, allows the learner to maintain the same motor task in short term memory without decay for much of the training which facilitates training performance, but hinders ability to retrieve the appropriate motor memory in a later testing context. A much more recent study [@chuaPracticeVariabilityPromotes2019], replicates the standard findings of an advantage of varied training over constant training (expt 1, bean-bag throwing task), and of random training over blocked training (expt 2 & 3, bean-bag throwing & golf putting). The novelty of this study is that the experimenters queried subjects about their attentional focus throughout the training stage. In all three experiments varied or random trained-subjects reported significantly greater external attention (e.g. attending to the target distance), and constant or blocked subjects reported more internal attention (e.g. posture or hand position). The authors argue that the benefits of varied/random training may be mediated by changes in attentional focus, however the claims made in the paper seem to go far beyond what can be justified by the analyses reported -- e.g. the increased external focus could be a simple byproduct of varied training. A stronger form of evidence that was not provided may have been to use multiple regression analyses to show that the testing advantage of the varied/random groups over the constant/blocked groups could be accounted for by the differences in self-reported attentional focus.

@@

(note that the studies described above manipulated the variability of the training examples, rather than the variability of the categories)

 Moreover, this effect was at odds with the predications of the baseline version of the GCM, thus providing some evidence that training variation may at least sometimes induce effects that cannot be entirely accounted for by exemplar-similarity accounts.

@@

Numerous researchers have attempted to provide coherent frameworks to account for the full range of influences of training variation and sequencing described above (along with many other effects not discussed). Such accounts are generally quite similar, invoking ideas of desirable levels of difficulty [@bjorkNewTheoryDisuse1992; @schmidtNewConceptualizationsPractice1992], or optimal challenge points [@guadagnoliChallengePointFramework2004]. They tend to start by describing the dissociation between acquisition performance (performance during training) and testing performance (delayed retention and/or transfer), most strikingly observed as varied/random training participants performing worse than their constant/blocked counterparts during the training stage of the study, but then outperforming the constant/blocked comparisons at a later retention or transfer stage. This observation is then used to justify the idea that the most enduring and generalizable learning occurs by training at an optimal level of training difficulty, with difficulty being some function of the experience of the learner, and the cognitive or visuomotor processing demands of the task. It then follows that the factors that tend to make training more difficult (i.e. increased variability or randomization), are more likely to be beneficial when the learner has some experience, or when the processing demands of the task are not too extreme (which may only occur after some experience with the task). Such frameworks may be helpful heuristics in some cases, but they also seem to be overly flexible such that any null result of some intervention might be accounted for by a suboptimal amount of training trials, or by suggesting the training task was too difficult. The development of computational models that can account for how changes in the parameters of the motor-skill task scale with difficulty, would be a great step forward.

; @wulfPrinciplesDerivedStudy2002

Through this combination of controlled experiments and computational modeling, the present work provides an extensive investigation into if, how, and when variability facilitates transfer

Together, these projects contribute to a deeper understanding of the role of training variability in cognitive development, offering insights that could inform educational and training practices across a variety of domains. The findings and methodological approaches detailed in this dissertation not only advance theoretical knowledge but also provide practical implications for designing more effective learning environments.

By taking a multi-pronged approach that combines empirical investigations across multiple experimental paradigms with computational modeling, this dissertation provides a comprehensive examination of the complex and sometimes opposing effects of training variability on learning and generalization. The contrasting results obtained from the HTT and HTW tasks underscore the importance of considering task characteristics when designing experiments to assess the influence of training manipulations. Furthermore, the computational modeling components of this work offer novel theoretical accounts for both the beneficial and detrimental effects of training variability observed in our experiments. Collectively, this dissertation highlights the value of integrating empirical and modeling approaches to uncover the cognitive mechanisms that support learning and generalization across a variety of domains.
